Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-11-14 pot/LaTeX/skewnormal.tex
plot updates and sim study tweak

The paper is organized as follows.
In section \ref{s:spatial}, we describe a spatial skew-$t$ process with partitioning and thresholding.
In section \ref{s:temporal}, we expand the model to incorporate temporal dependence.
The computing is described in section \ref{s:comp}.
In section \ref{s:simstudy}, we present a simulation study that examines the predictive capabilities of this model compared with a na{\"i}ve Gaussian method.
We then compare our method to Gaussian and max-stable methods with a data analysis of ozone measurements from the eastern US in section \ref{s:analysis}.
The final section provides brief discussion and direction for future research.

\section{Spatial model}\label{s:spatial}
To avoid bias in estimating tail parameters, we model censored data.
Let
\beq\label{Yt}
  \tilde{Y}(\bs) = \delta(\bs) Y(\bs) + (1 - \delta(\bs)) T
\eeq
be the censored observation at site $\bs$ where $Y(\bs)$ is the uncensored observation, $\delta(\bs) = I[Y(\bs) > T]$, and $T$ is a pre-specified threshold value.
Then, assuming the uncensored data are observations from a skew-$t$ process, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep[see Appendix A.5]{Azzalini2003}.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  \bY &= \bX \bbeta + \alpha |z| + \sigma \bv
\end{align}
where $\alpha \in \calR$ controls the skewness; $z \sim N(0, \sigma^2)$ is a random effect from a half-normal distribution (see appendix A.4); $\bv$ is a Gaussian process with mean zero, variance one, \Matern spatial correlation with spatial range $\rho$, smoothness $\nu$, and $\gamma$ is the proportion of variance accounted for by the spatial variation; and $\sigma^2 \sim \text{IG}(a, b)$.
When marginalizing over the $z$ and $\sigma^2$ terms, the joint distribution for all sites becomes
\begin{align*}
  \bY \sim \text{skew-}t (\mu, \bSigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\bSigma^* = \frac{ b }{ a } \bSigma$, $\bSigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

\subsection{Extremal dependence}
One measure of extremal dependence is the $\chi$ statistic \citep{Padoan2011}.
The $\chi$ statistic for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c).
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(h) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(h) > c)
\end{align*}
where $h = ||\bs_1 - \bs_2||$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the skew-$t$ process described in \ref{s:data}, $\chi(h) > 0$ \citep[they give an expression that is the sum of two survival functions from the skew-$t$ distribution. The expression isn't terribly informative, but I can include it if you think it would be helpful.]{Padoan2011}.

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z$ and $\sigma^2$ terms.
Even more problematic is that $\lim_{h \rightarrow \infty} \chi(h) \neq 0$.
We handle this problem with a random partition similar to \citet{Kim2005} that allows $z$ and $\sigma^2$ to vary by site.

@@

\section{Accounting for temporal dependence} \label{s:temporal}
In a standard block-maxima approach, the temporal dependence is assumed to be negligible because blocks (e.g. yearly maxima) are taken to be large enough to assume independence.
However, when using daily meausrements, the assumption of temporal independence is no longer appropriate.
There are several places where temporal dependence could be included, including the residual $v(\bs)$.
However, we choose to allow for temporal dependence in the $\bw$, $z(\bs)$, and $\sigma(\bs)$ terms because these terms derive the tail behavior.
Specifically, we incorporate an AR(1) time series.
We first, transform the spatial knots from $\calD$ to $\calR^2$ as follows.
Let
\begin{align*}
  \bw^*_k = \Phi^{-1}_2\left[ \frac{ (\bw_k - \min(\bs))}{ \text{range}(\bs) } \right].
\end{align*}
\hl{What's the best way to write the $\bw$ transformation as 1-D? $w_x$ and $w_y$ or $w_1$ and $w_2$?}
where $\Phi_2$ represents a bivariate standard normal density function.
Then $\bw^*_k \in \calR^2$.
We use a copula on the $\sigma^2_t(\bs)$ to ensure that the marginal distributions of $\sigma^2_t(\bs)$ are inverse gamma.
Let
\begin{align*}
  \sigma^{2*}_t(\bs) =\Phi^{-1}\left\{ \text{IG}[\sigma^2_t(\bs)] \right\}
\end{align*}
where $\Phi$ is a univariate standard normal density function, and IG is the distribution function for an IG$(a, b)$ random variable.
Then, when $t > 1$ the time series is modeled as
\begin{align*}
  \bw^*_{1, k} & \sim N_2(0, 1)\\
  z_{1k} & \sim N(0, \sigma^2_{1k})\\
  \sigma^{2*}_{1k} &\sim N(0, 1)\\
  \bw^*_{tk} | \bw^*_{t-1, k} &\sim N_2\left[\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2) \right] \\
  z_{tk} | z_{t-1, k} &\sim N \left[\phi_z z_{t-1, k}, \sigma^2_{tk} (1 - \phi_z^2)\right] \\
  \sigma^{2*}_{tk} | \sigma^{2*}_{t-1, k} &\sim N \left[\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2) \right]
\end{align*}
where $|\phi_w| < 1$, $0 < \phi_z < 1$, and $|\phi_\sigma| < 1$.
These are stationary time series models with marginal distributions
\begin{align*}
  \bw^*_{k} & \sim N_2(0, 1)\\
  z_{k} & \sim N(0, \sigma^2_{k})\\
  \sigma^{2*}_{k} &\sim N(0, 1).
\end{align*}
For each day, the model is identical to the spatial-only model in Seciton \ref{s:part}.

>>>>

The paper is organized as follows.
Section \ref{s:skewt} is a brief review of the spatial skew-$t$ process.
In section \ref{s:temporal}, we build upon the traditional skew-$t$ by incorporating censoring to focus on tails, partitioning to remove long-range asymptotic dependence, and extending the model to space-time data.
The computing is described in section \ref{s:comp}.
In section \ref{s:simstudy}, we present a simulation study that examines the predictive capabilities of this model compared with a na{\"i}ve Gaussian method.
We then compare our method to Gaussian and max-stable methods with a data analysis of ozone measurements from the eastern US in section \ref{s:analysis}.
The final section provides brief discussion and direction for future research.

\section{Spatial skew processes}\label{s:skewt}
Many types of data demonstrate some level of skewness and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing a slant parameter $\alpha$ to account for non-symmetric data \citep{Genton2004}.

\subsection{Skew-normal process}
Let $U(\bs)$ be a normalized skew-normal process with slant $\alpha$. We write $U(\bs) \sim \text{SN}_d(\Omega, \alpha)$, and the density is given as
\begin{align}
  2 \phi_d (U(\bs); \Omega) \Phi(\alpha \bOne^T U(\bs))
\end{align}
where $\phi_d$ is the density function for a $N_d(0, \Sigma)$ random variable, and $\Phi$ is a standard normal cumulative distribution function.
There is an additive representation \citep{Zhang2010} to construct a standardized skew-normal process given by
\begin{align}
  U(\bs) = \alpha |z| + v(\bs)
\end{align}
where $\alpha \in \calR$ is a parameter controlling skew, $z \sim N(0, 1)$, and $v(\bs)$ is a Gaussian process with mean zero, variance one, \Matern spatial correlation with spatial range $\rho$, smoothness $\nu$, and $\gamma$ is the proportion of variance accounted for by the spatial variation.
So,
\begin{align}
  \text{cor}(U(\bs), U(\bs)) = \gamma I(s = t) + (1 - \gamma)  \frac{ 1 }{ \Gamma(\nu) 2^{ \nu - 1}} \left( \sqrt{2\nu} \frac{ ||\bs - \bt|| }{ \rho } \right)^{\nu} K_{\nu} \left( \sqrt{2\nu} \frac{ ||\bs - \bt|| }{ \rho } \right).
\end{align}
Let $Y(\bs) = \bX(\bs)^T + \sigma U(\bs)$.
Then, after marginalizing over the $z$ term, we find that
\begin{align}
  Y(\bs) \sim \text{SN}_d(\bX^T(\bs)\bbeta, \Omega, \alpha)
\end{align}
where $\Omega = \sigma^2 (\Sigma + \alpha^2 \bOne \bOne^T$).

\subsection{Skew-$t$ process}
If $\sigma^2 \sim \text{IG}(a, b)$, after over the $\sigma^2$ terms, we have a skew-$t$ process with distribution function

\begin{align*}
  \bY \sim \text{skew-}t (\mu, \bSigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\bSigma^* = \frac{ b }{ a } \bSigma$, $\bSigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

One member of this family, is the skew-$t$ distribution \citep[see Appendix A.5]{Azzalini2003}.
\citet{Zhang2010} show that the skew-$t$ process can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y(\bs) &= \bX(\bs)^T \bbeta + \alpha |z| + \sigma v(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness; $z \sim N(0, \sigma^2)$ is a normal random effect; $v(\bs)$ is a Gaussian process with mean zero, variance one, \Matern spatial correlation with spatial range $\rho$, smoothness $\nu$, and $\gamma$ is the proportion of variance accounted for by the spatial variation; and $\sigma^2 \sim \text{IG}(a, b)$.
So,
\begin{align}
  \text{cor}(Y(\bs), Y(\bs)) = \gamma I(s = t) + (1 - \gamma)  \frac{ 1 }{ \Gamma(\nu) 2^{ \nu - 1}} \left( \sqrt{2\nu} \frac{ ||\bs - \bt|| }{ \rho } \right)^{\nu} K_{\nu} \left( \sqrt{2\nu} \frac{ ||\bs - \bt|| }{ \rho } \right)
\end{align}

\subsection{Extremal dependence}
One measure of extremal dependence is the $\chi$ statistic \citep{Padoan2011}.
The $\chi$ statistic for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c).
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(h) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(h) > c)
\end{align*}
where $h = ||\bs_1 - \bs_2||$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the skew-$t$ process described in \ref{s:skewt}, $\chi(h) > 0$ \citep[they give an expression that is the sum of two survival functions from the skew-$t$ distribution. The expression isn't terribly informative, but I can include it if you think it would be helpful.]{Padoan2011}.

\section{Spatiotemporal skew-$t$ model for extremes}\label{s:spatial}
In this section, we describe the

\subsection{Censoring to focus on the tails}
To avoid bias in estimating tail parameters, we model censored data.
Let
\beq\label{Yt}
  \tilde{Y}(\bs) = \left\{ \begin{array}{ll}
      Y(\bs) \quad & \delta(\bs) = 1 \\
      T & \delta(\bs) = 0
  \end{array} \right.
\eeq
be the censored observation at site $\bs$ where $Y(\bs)$ is the uncensored observation, $\delta(\bs) = I[Y(\bs) > T]$, and $T$ is a pre-specified threshold value.
Then, assuming the uncensored data $Y(\bs)$ are observations from a skew-$t$ process, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Partitioning to remove long-range asymptotic dependence}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z$ and $\sigma^2$ terms.
Even more problematic is that $\lim_{h \rightarrow \infty} \chi(h) \neq 0$.
We handle this problem with a random partition similar to \citet{Kim2005} that allows $z$ and $\sigma^2$ to vary by site.

@@

\subsection{Extension to space-time data} \label{s:temporal}
In a standard block-maxima approach, the temporal dependence is assumed to be negligible because blocks (e.g. yearly maxima) are taken to be large enough to assume independence.
However, when using daily meausrements, the assumption of temporal independence is no longer appropriate.
There are several places where temporal dependence could be included, including the residual $v(\bs)$.
However, we choose to allow for temporal dependence in the $\bw$, $z(\bs)$, and $\sigma(\bs)$ terms because these terms derive the tail behavior.
Specifically, we incorporate an AR(1) time series.
We first, transform the spatial knots from $\calD$ to $\calR^2$ as follows.
Let
\begin{align*}
  \bw^*_k = \Phi^{-1}_2\left[ \frac{ (\bw_k - \min(\bs))}{ \text{range}(\bs) } \right].
\end{align*}
\hl{What's the best way to write the $\bw$ transformation as 1-D? $w_x$ and $w_y$ or $w_1$ and $w_2$?}
where $\Phi_2$ represents a bivariate standard normal density function.
Then $\bw^*_k \in \calR^2$.
We use a copula on the $\sigma^2_t(\bs)$ to ensure that the marginal distributions of $\sigma^2_t(\bs)$ are inverse gamma.
Let
\begin{align*}
  \sigma^{2*}_t(\bs) =\Phi^{-1}\left\{ \text{IG}[\sigma^2_t(\bs)] \right\}
\end{align*}
where $\Phi$ is a univariate standard normal density function, and IG is the distribution function for an IG$(a, b)$ random variable.
Then, when $t > 1$ the time series is modeled as
\begin{align*}
  \bw^*_{1, k} & \sim N_2(0, 1)\\
  z_{1k} & \sim N(0, \sigma^2_{1k})\\
  \sigma^{2*}_{1k} &\sim N(0, 1)\\
  \bw^*_{tk} | \bw^*_{t-1, k} &\sim N_2\left[\phi_w \bw^*_{t-1, k}, (1 - \phi_w^2) \right] \\
  z_{tk} | z_{t-1, k} &\sim N \left[\phi_z z_{t-1, k}, \sigma^2_{tk} (1 - \phi_z^2)\right] \\
  \sigma^{2*}_{tk} | \sigma^{2*}_{t-1, k} &\sim N \left[\phi_\sigma \sigma^{2*}_{t-1, k}, (1 - \phi_\sigma^2) \right]
\end{align*}
where $|\phi_w| < 1$, $0 < \phi_z < 1$, and $|\phi_\sigma| < 1$.
These are stationary time series models with marginal distributions
\begin{align*}
  \bw^*_{k} & \sim N_2(0, 1)\\
  z_{k} & \sim N(0, \sigma^2_{k})\\
  \sigma^{2*}_{k} &\sim N(0, 1).
\end{align*}
For each day, the model is identical to the spatial-only model in Seciton \ref{s:part}.
